---
title: "stanmodels"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(1)
```

```{r}
library(rstan)
library(loo)
```

# Pooled Model

```{r}
data <- read.csv("../data/preprocessed_data.csv")
data
```

```{r}
x1 <- data[,'Continent']
x2 <- data[, 'Fullscale_IQ_Change']
x3 <- data[,'Schooling_Index']
x4 <- data[,'Decade']
```

## Model with Decade Index

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           y =  x2,
           xpred = 11,
           seed = 1)
fit <- stan(file = '../models/pooled/pool.stan', data = IQ)

m <- monitor(fit)
```

```{r}
ans <- extract(fit)
mean(ans$a)
mean(ans$b)
hist(ans$a)
hist(ans$b)
```

```{r}
mean(ans$ypred)
quantile(ans$ypred, c(0.05, 0.1,0.5, 0.9, 0.95))
```

Convergence Analysis

```{r}
fit
check_hmc_diagnostics(fit)
```

```{r}
print("a")
m$Rhat[1]
m$n_eff[1]
print("---")
print("b")
m$Rhat[2]
m$n_eff[2]
```

```{r}
log_lik_1 <- extract_log_lik(fit, merge_chains = FALSE)
r_eff <- relative_eff(exp(log_lik_1))
loo_1 <- loo(log_lik_1, r_eff = r_eff)
loo_1$estimates
```

```{r}
a <- mean(ans$a)
b <- mean(ans$b)
alpha <- quantile(ans$a, c(0.05, 0.95))
beta <- quantile(ans$b, c(0.05, 0.95))
##png(file="plot.png")
plot(x = x4, 
     y = a+b*x4, 
     type='l', 
     col='blue',
     xlab ='Decade Index',
     ylab = 'Full Scale IQ Change')
points(x4, x2)
lines(x4, alpha[1] + beta[1] *x4, col ='red', lty=3)
lines(x4, alpha[2] + beta[2] *x4, col ='red', lty=3)
```

## Model with Schooling Index

```{r}
IQ <- list(N = length(x3),
           x1 = x3,
           y = x2)
fit <- stan(file = '../models/pooled/pool_schooling.stan', data = IQ, seed=1)
m <- monitor(fit)
```

```{r}
ans <- extract(fit)
mean(ans$a)
mean(ans$c)
hist(ans$a)
hist(ans$c)
```

Convergence Diagnostics

```{r}
fit
check_hmc_diagnostics(fit)
```

```{r}
print("a")
m$Rhat[1]
m$n_eff[1]
print("---")
print("c")
m$Rhat[2]
m$n_eff[2]
```

```{r}
log_lik_1 <- extract_log_lik(fit, merge_chains = FALSE)
r_eff <- relative_eff(exp(log_lik_1))
loo_1 <- loo(log_lik_1, r_eff = r_eff)
loo_1$estimates
```

## Model with Decade and Schooling Index

```{r}
latest_data <- data[data['Decade'] == 10]
n <- length(latest_data)
schooling <- as.numeric(unlist(latest_data[16:n]))
mu <- mean(schooling)
mu_vec <- c(mu * 1.05, mu * 1.1, mu * 1.15)
```

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           x2 = x3,
           y = x2,
           xpred_decade = 11,
           xpred_schooling = mu_vec
)
fit <- stan(file = '../models/pooled/pool_adv.stan', data = IQ, seed=1)
m <- monitor(fit)
```

```{r}
ans <- extract(fit)
hist(ans$a)
mean(ans$a)
hist(ans$b)
mean(ans$b)
hist(ans$c)
```

Prediction

```{r}
mean(ans$ypred[,1])
mean(ans$ypred[,2])
mean(ans$ypred[,3])
```

Convergence Diagnostics

```{r}
fit
check_hmc_diagnostics(fit)
```

```{r}
print("a")
m$Rhat[1]
m$n_eff[1]
print("---")
print("b")
m$Rhat[2]
m$n_eff[2]
print("c")
m$Rhat[3]
m$n_eff[3]
```

```{r}
log_lik_1 <- extract_log_lik(fit, merge_chains = FALSE)
r_eff <- relative_eff(exp(log_lik_1))
loo_1 <- loo(log_lik_1, r_eff = r_eff)
loo_1$estimates
```

```{r}
mean(ans$ypred)
quantile(ans$ypred, c(0.05, 0.1,0.5, 0.9, 0.95))
```

```{r}
a <- mean(ans$a)
b <- mean(ans$b)
q_a <- quantile(ans$a, c(0.05, 0.95))
q_b <- quantile(ans$b, c(0.05, 0.95))

plot(x=x4, y=x3, type='p', ylim = c(-0.25,1))
lines(x4, a + b*x4, col ="red")
```

```{r}
a <- mean(ans$a)
b <- mean(ans$b)
q_a <- quantile(ans$a, c(0.05, 0.95))
q_b <- quantile(ans$b, c(0.05, 0.95))

plot(x=x4, y=x3, type='p', ylim=c(-6,6))
lines(x4, a + b*x4, col ="red")
lines(x4, q_a[1] + q_b[1]*x4, lty=3)
lines(x4, q_a[2] + q_b[2]*x4, col="blue", lty=3)
```

```{r}
dim(ans$b)
```

Prior Sensitivity Analysis

Simple Model

Default Priors

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           y =  x2)
fit <- stan(file = '../models/pooled/pool.stan', data = IQ, seed=1)
m <- monitor(fit)
```

```{r}
ans <- extract(fit)
mean(ans$a)
mean(ans$b)
hist(ans$a)
hist(ans$b)
hist(ans$sigma)
mean(ans$sigma)
```

a set to Normal(0,10)

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           y =  x2)
fit <- stan(file = '../models/pooled/pool_prior1.stan', data = IQ, seed=1)
m <- monitor(fit)
```

```{r}
ans <- extract(fit)
mean(ans$a)
mean(ans$b)
hist(ans$a)
hist(ans$b)
hist(ans$sigma)
mean(ans$sigma)
```

Sigma set to inv_chi_sq(1)

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           y =  x2)
fit <- stan(file = '../models/pooled/pool.stan', data = IQ, seed=1)
m <- monitor(fit)
```

```{r}
ans <- extract(fit)
mean(ans$a)
mean(ans$b)
hist(ans$a)
hist(ans$b)
hist(ans$sigma)
mean(ans$sigma)
```

Analysis: *a*'s distribution, when the prior is changed from N(0,1) to N(0,10) remains a normal distribution but shifts its center from 0.86 to 2.4. This implies that our model is sensitive to the prior distribution of a.

While sigma's distribution stays almost identical when changed from N(0,100) to inv_chi_square(1). This implies that our model is insensitive to the prior distribution of sigma.

------------------------------------------------------------------------

Advanced Pooled model

Default Priors

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           x2 = x3,
           y = x2,
           xpred_decade = 11,
           xpred_schooling = mu_vec
)
fit <- stan(file = '../models/pooled/pool_adv.stan', data = IQ, seed=1)
fit_prior <- stan(file = '../models/pooled/pool_adv_prior1.stan', data = IQ, seed=1)
```

```{r}
ans_default <- extract(fit)
ans_prior1 <- extract(fit_prior)

hist( ans_default$a, col="lightblue", xlim=c(-5,10), xlab="a")  # first histogram
hist( ans_prior1$a, col=rgb(1,0,0,1/4), xlim=c(-5,10), add=T)  # second
legend(x="topright", legend=c("Normal Prior", "Modified Prior"), fill=c("lightblue", rgb(1,0,0,1/4)))
```

a set to N(0, 10)

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           x2 = x3,
           y = x2,
           xpred_decade = 11,
           xpred_schooling = mu_vec
)
fit <- stan(file = '../models/pooled/pool_adv_prior1.stan', data = IQ, seed=1)
m <- monitor(fit)
```

```{r}
ans <- extract(fit)
mean(ans$a)
mean(ans$b)
hist(ans$a)
hist(ans$b)
hist(ans$sigma)
mean(ans$sigma)
```

Sigma set to inv_chi_square(1)

```{r}
IQ <- list(N = length(x3),
           x1 = x4,
           x2 = x3,
           y = x2,
           xpred_decade = 11,
           xpred_schooling = mu_vec
)
fit <- stan(file = '../models/pooled/pool_adv_prior2.stan', data = IQ, seed=1)
m <- monitor(fit)
```

```{r}
ans <- extract(fit)
mean(ans$a)
mean(ans$b)
hist(ans$a)
hist(ans$b)
hist(ans$sigma)
mean(ans$sigma)
```
